The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2013-05-14 script/en/chapters/numbertheory.tex
CT Script: Minor changes.

\newpage
\hypertarget{Chapter_ElementaryNT}{}
\chapter{Introduction to Elementary Number Theory with Examples}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, July 2001; Updates: Dec. 2001, June 2002, May 2003,
 May 2005, March 2006, June 2007, July 2009, Jan. 2010, Dec. 2012) \\

@@

What do the numbers $5, 12, 19, 26, \cdots$ have in common for division by $7$? The remainder is always $r = 5$.
For division by $7$, only the following remainders are possible:
$$r = 0, 1, 2, \cdots, 6.$$

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Calculations with finite sets}

@@

From algebra theorems it follows that essential parts of the conventional
calculation rules are kept when we proceed to modular calculations over
a basic set $\mathbb{Z}$.  For example, addition remains commutative. 
The same goes for multiplication modulo $m$. The result of a
division\footnote{%
The division modulo $m$\index{Division modulo $n$} is only defined for
numbers co-prime\index{Number!co-prime} to $m$ because other numbers
have the same property as zero (this means there is no inverse number).
See footnote \ref{ftn-mod6} in chapter~\ref{addmult} and
table~\ref{mulmod6} in chapter~\ref{add-and-mult-inverses}.
\label{ftn-res-divmodn}\label{ftn-zth-divmodn}
} is not a fraction but rather a whole number between $0$ and $m-1$.

The known laws apply:
\begin{itemize}
\item[\bf 1.] {\bf Associative law:}\index{Associative law} \\ 
    $((a+b) + c) {\rm ~(mod~ } m) \equiv  (a + (b+c)) {\rm ~(mod~ } m).$ \\
    $((a*b) * c) {\rm ~(mod~ } m) \equiv  (a * (b*c)) {\rm ~(mod~ } m).$
\item[\bf 2.] {\bf Commutative law:} \index{Commutative law}\\
    $(a+b) {\rm ~(mod~ } m) \equiv  (b+a) {\rm ~(mod~ } m).$ \\
     $(a*b) {\rm ~(mod~ } m) \equiv  (b*a) {\rm ~(mod~ } m).$
\end{itemize}
The associative law and the commutative law apply to both addition and multiplication.
\begin{itemize}
\item[\bf 3.] {\bf Distributive law:} \index{Distributive law}\\
    $ (a * (b+c)) {\rm ~(mod~ } m) \equiv  (a*b + a*c) {\rm ~(mod~ } m).$ 
\item[\bf 4.] {\bf Reducibility:} \index{Reducibility} \\
    $(a+b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) + b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\  
    $(a*b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) * b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\
    When adding or multiplying the order in which the modulo operation is performed does not matter.
\end{itemize}
\begin{itemize}
\item[\bf 5.] {\bf Existence of an identity (neutral element):} \index{Identity}\\
    $(a + 0) {\rm ~(mod~ } m) \equiv  (0 + a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$  \\
    $(a * 1) {\rm ~(mod~ } m) \equiv  (1 * a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$
\item[\bf 6.] {\bf Existence of an inverse element:} \\
    For all whole numbers $a$ and $m$ there exists a whole number $-a$ such that: \\
    $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$ \quad (additive inverse).
\index{Inverse!additive}\\
    For each $a$ ($a \not\equiv 0 {\rm ~(mod~ } p$) ) where $p$ is prime there exists a whole number $a^{-1}$, 
such that: \\
    $(a * a^{-1}) {\rm ~(mod~ } p) \equiv 1 {\rm ~(mod~}p)$ \quad (multiplicative inverse). 
\index{Inverse!multiplicative}
\item[\bf 7.] \index{Closeness} {\bf Closeness}\footnote{\label{ftn-zth-closed}The
   property of closeness is always defined in relation to an operation in a set.
   See chapter~\ref{nt:NumberTheory_Appendix_B} ``\nameref{nt:NumberTheory_Appendix_B}''.}: \\
      %%% \hyperlink{nt:NumberTheory_Appendix_B}{Appendix B of this chapter}.}:   \\
     $a, b \in G  \Longrightarrow  ( a + b ) \in G.$ \\
    $a, b \in G  \Longrightarrow  ( a * b ) \in G.$
\item[\bf 8.] \index{Transitivity} {\bf Transitivity}:

>>>>

\newpage
\hypertarget{Chapter_ElementaryNT}{}
\chapter{Introduction to Elementary Number Theory with Examples}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, July 2001; Updates: Dec. 2001, June 2002, May 2003,
 May 2005, March 2006, June 2007, July 2009, Jan. 2010, May 2013) \\

@@

What do the numbers $5, 12, 19, 26, \cdots$ have in common for division by $7$? The remainder is always $r = 5$.
For division by $7$, only the following remainders are possible:
$$r = 0, 1, 2, \cdots, 6$$

If $r = 0$, then: $m | n$ (``$m$ divides $n$'').

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\section{Calculations with finite sets}

@@

From algebra theorems it follows that essential parts of the conventional
calculation rules are kept when we proceed to modular calculations over
a basic set $\mathbb{Z}$.  For example, addition remains commutative. 
The same goes for multiplication modulo $m$. The result of a
division\footnote{%
The division modulo $m$\index{Division modulo $n$} is only defined for
numbers co-prime\index{Number!co-prime} to $m$ because other numbers
have the same property as zero (this means there is no inverse number).
See law number 6 {\bf existence of an inverse element}.
See footnote \ref{ftn-mod6} in chapter~\ref{addmult} and
table~\ref{mulmod6} in chapter~\ref{add-and-mult-inverses}.
\label{ftn-res-divmodn}\label{ftn-zth-divmodn}
} is not a fraction but rather a whole number between $0$ and $m-1$.

The known laws apply:
\begin{itemize}
\item[\bf 1.] {\bf Associative law:}\index{Associative law} \\ 
    $((a+b) + c) {\rm ~(mod~ } m) \equiv  (a + (b+c)) {\rm ~(mod~ } m).$ \\
    $((a*b) * c) {\rm ~(mod~ } m) \equiv  (a * (b*c)) {\rm ~(mod~ } m).$
\item[\bf 2.] {\bf Commutative law:} \index{Commutative law}\\
    $(a+b) {\rm ~(mod~ } m) \equiv  (b+a) {\rm ~(mod~ } m).$ \\
     $(a*b) {\rm ~(mod~ } m) \equiv  (b*a) {\rm ~(mod~ } m).$
\end{itemize}
The associative law and the commutative law apply to both addition and multiplication.
\begin{itemize}
\item[\bf 3.] {\bf Distributive law:} \index{Distributive law}\\
    $ (a * (b+c)) {\rm ~(mod~ } m) \equiv  (a*b + a*c) {\rm ~(mod~ } m).$ 
\item[\bf 4.] {\bf Reducibility:} \index{Reducibility} \\
    $(a+b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) + b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\  
    $(a*b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) * b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\
    When adding or multiplying the order in which the modulo operation is performed does not matter.
\end{itemize}

\begin{itemize}
\item[\bf 5.] {\bf Existence of an identity (neutral element):} \index{Identity}\\
    $(a + 0) {\rm ~(mod~ } m) \equiv  (0 + a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$  \\
    $(a * 1) {\rm ~(mod~ } m) \equiv  (1 * a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$

\item[\bf 6.] {\bf Existence of an inverse element}\footnote{%
An inverse element only exits, if it is unique for the given operation.
}:
% \begin{itemize}
\begin{compactitem}

\item {\bf Additive inverse}\index{Inverse!additive}\\
   For all whole numbers $a$ and $m$ there exists a whole number $-a$ such
   that: \\
   $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$
   % $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$ \quad (additive inverse)

\item {\bf Multiplicative inverse modulo a prime p}\\
    For each whole number $a$ (with $a \not\equiv 0 {\rm ~(mod~ } p$) and $p$
    prime) there exists a whole number $a^{-1}$, such that:~~~ 
    $(a * a^{-1}) {\rm ~(mod~ } p) \equiv 1 {\rm ~(mod~}p)$
    % \quad (multiplicative inverse)
    \index{Inverse!multiplicative}

\item {\bf Multiplicative inverse modulo a compound number m}\footnote{%
    As $8 \equiv 3 ~mod~ 5$ and $ 3 * 2 \equiv 1 ~mod~ 5$,
    then $ 2 = 3^{-1} = 8^{-1}$ is a (unique) inverse for $3$ and $8$.\\
    A multiple of p or m has no inverse mod p or mod m:
    $5 \equiv 10 \equiv 0 ~mod~ 5$.
}\\
    For all whole numbers $a$ and $m$ (with $a \not\equiv 0 {\rm ~(mod~ } m$)
    and $ggT(a,m) = 1$) there exists a whole number $a^{-1}$, such that: 
    $(a * a^{-1}) {\rm ~(mod~ } m) \equiv 1 {\rm ~(mod~}m)$

% \end{itemize}
\end{compactitem}

\item[\bf 7.] \index{Closeness} {\bf Closeness}\footnote{\label{ftn-zth-closed}The
   property of closeness is always defined in relation to an operation in a set.
   See chapter~\ref{nt:NumberTheory_Appendix_B} ``\nameref{nt:NumberTheory_Appendix_B}''.}: \\
      %%% \hyperlink{nt:NumberTheory_Appendix_B}{Appendix B of this chapter}.}:   \\
     $a, b \in G  \Longrightarrow  ( a + b ) \in G.$ \\
    $a, b \in G  \Longrightarrow  ( a * b ) \in G.$

\item[\bf 8.] \index{Transitivity} {\bf Transitivity}:

### 2013-05-14 script/en/chapters/paper_and_pencil.tex
CT Script: Minor changes.

\chapter{Paper and Pencil Encryption Methods}
\label{Kapitel_PaperandPencil}
(Christine St\"otzel, April 2004; Updates: B.+C. Esslinger, June 2005;
 Updates Minh Van Nguyen and B. Esslinger, Nov. 2009, June 2010, Jan. 2013)
\index{Paper and pencil methods}
\index{Encryption!classic}
\index{Cryptography!classic}

@@

   Unformated Plaintext: plaintext letters are encrypted in pairs\\
   Formated Plaintext:   pl ai nt ex tl et te rs ar ee ncrypted in pairs\\
   Formated Plaintext:   plaintext letters are x encrypted in pairs\\
   Formated Plaintext:   pl ai nt ex tl et te rs ar ex en cr yp te di np ai rs

@@

\item {\bf Card games} \cite{pp:Savard1999}: The key is created by means of a
   pack of cards and rules that are agreed upon in advance. All methods
   mentioned in this paragraph are designed as paper and pencil methods,
   i.e. they are applicable without electronic aid. A pack of cards is
   unsuspicious to outsiders, shuffling the deck provides a certain amount
   of coincidence, cards can be transformed into numbers easily and a
   transposition cipher can be carried out without any further aid.
   \begin{itemize}
      \item {\bf Solitaire cipher (Bruce Schneier)\footnotemark}
         \footnotetext{%
         In CT1\index{CrypTool 1} you can call this method under the menu
         {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
         Solitaire}.}
         \cite{pp:Schneier1999}\index{Solitaire}: 
         Sender and receiver have to own a deck of cards shuffled in the
         same manner. A key stream is generated that has to consist of as 
         many characters as the message to be encrypted.

@@

The Hill~\cite{pp:Hill1929,pp:Hill1931} or matrix cipher is more
mathematically sophisticated than other ciphers mentioned in this
chapter. The encryption/decryption key of this cipher is an invertible
square matrix, and the plaintext/ciphertext is processed also as a matrix.
The encryption and decryption processes use matrix operations modulo 26.
The Hill cipher is implemented in the Sage class
\begin{center}
\verb!sage.crypto.classical.HillCryptosystem!
\end{center}

\noindent In the following example our plaintext/ciphertext space is the
capital letters of the English alphabet. In the Hill cipher, each
letter of this alphabet is assigned a unique integer modulo 26.
The size of matrix (also called its dimension) is not restricted by the cipher.

\noindent {\bf Remark:} Comparing the Hill implementation in CrypTool v1.4.31
and in Sage v5.3:
\begin{itemize}
  \item Sage offers fast commandline operations; CT1 offers
        its functionality within a GUI.
  \item Sage offers any dimension; CT1 is restricted
        to a matrix size between 1 and 10.
  \item Sage allows negative numbers in the key matrix, and converts them
        automatically into appropriate non-negative numbers;
        CT1 doesn't allow negative numbers in the key matrix.
  \item Sage always sets the first alphabet character to 0;\\
        Sage only allows the 26 capital letters as alphabet;\\
        and it uses only the multiplication variant
        plaintext row vector * key matrix:\\
        C = P*M.
  \item CT1 offers to choose also 1 as value for the first alphabet
        character; you can combine your alphabet within the text options dialog;
        and it offers to use a reverse multiplication variant too.
\end{itemize}

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.45]{figures/PaP_Fig_Hill-Cipher-CT-Dialog.png}
\caption{Hill dialog in CT1 with the operations and options available\vspace{1ex}} 
\label{PaP_Fig_Hill-Cipher-CT-Dialog}
\end{center}
\end{figure}

>>>>

\chapter{Paper and Pencil Encryption Methods}
\label{Kapitel_PaperandPencil}
(Christine St\"otzel, April 2004; Updates: B.+C. Esslinger, June 2005;
 Updates Minh Van Nguyen and B. Esslinger, Nov. 2009, June 2010, May 2013)
\index{Paper and pencil methods}
\index{Encryption!classic}
\index{Cryptography!classic}

@@

   Unformated Plaintext: plaintext letters are encrypted in pairs\\
   Formated Plaintext:   pl ai nt ex tl et te rs ar ee ncrypted in pairs\\
   Formated Plaintext:   pl ai nt ex tl et te rs ar ex en cr yp te di np ai rs

@@

\item {\bf Card games} \cite{pp:Savard1999}: The key is created by means of a
   pack of cards and rules that are agreed upon in advance. All methods
   mentioned in this paragraph are designed as paper and pencil methods,
   i.e. they are applicable without electronic aid. A pack of cards is
   unsuspicious to outsiders, shuffling the deck provides a certain amount
   of coincidence, cards can be transformed into numbers easily and a
   transposition cipher can be carried out without any further aid.
   \begin{itemize}
      \item {\bf Solitaire cipher (Bruce Schneier)\footnotemark}
         \footnotetext{%
         In CT1\index{CrypTool 1} you can call this method under the menu
         {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
         Solitaire}.\\
          In CT2\index{CrypTool 2} you can find this method within the templates
          {\bf Cryptography \textbackslash{} Classical} and 
          {\bf Cryptanalysis \textbackslash{} Classical}.
         }
         \cite{pp:Schneier1999}\index{Solitaire}: 
         Sender and receiver have to own a deck of cards shuffled in the
         same manner. A key stream is generated that has to consist of as 
         many characters as the message to be encrypted.

@@

The Hill~\cite{pp:Hill1929,pp:Hill1931} or matrix cipher is
mathematically more sophisticated than the other ciphers mentioned in this
chapter. The encryption/decryption key of this cipher is an invertible
square matrix (here called $key$).
Plaintext and ciphertext are vectors ($P$ and $C$).
The encryption and decryption processes use matrix operations modulo 26,
here it is $C = P * key$ $(mod$ $26)$. %Zweimal $$-Klammer, damit Blank vor "(".
The Hill cipher is implemented in the Sage class
\begin{center}
\verb!sage.crypto.classical.HillCryptosystem!
\end{center}

\noindent In the following example our plaintext/ciphertext space is the
capital letters of the English alphabet. In the Hill cipher, each
letter of this alphabet is assigned a unique integer modulo 26.
The size of the key matrix (also called its dimension) is not restricted by the
cipher.\\

\noindent {\bf Remark:} Comparing the Hill implementation in CrypTool v1.4.31
and in Sage version 5.3:
\begin{itemize}
  \item Sage offers fast commandline operations; CT1 offers
        its functionality within a GUI.
  \item Sage offers for the key matrix any dimension; CT1 is restricted
        to a matrix size between 1 and 10.
  \item Sage allows negative numbers in the key matrix, and converts them
        automatically into appropriate non-negative numbers;
        CT1 doesn't allow negative numbers in the key matrix.
  \item Sage always sets the first alphabet character to 0;\\
        Sage only allows the 26 capital letters as alphabet;\\
        and it uses only the multiplication variant
        plaintext row vector * key matrix:\\
        $C = P * key$.
  \item CT1 offers to choose also 1 as value for the first alphabet
        character; you can combine your alphabet within the text options dialog;
        and it also offers to use a reverse multiplication variant:
        $C = key * P$.
\end{itemize}

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.7]{figures/PaP_Fig_Hill-Cipher-CT-Dialog.png}
\caption{Hill dialog in CT1 with the operations and options available\vspace{1ex}} 
\label{PaP_Fig_Hill-Cipher-CT-Dialog}
\end{center}
\end{figure}

### 2013-05-14 script/en/chapters/primes.tex
CT Script: Minor changes.

\newpage
% \hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_Primes}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, Jan. 2010, Feb. 2013)

@@

% --------------------------------------------------------------------------
\section{What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers (composite
numbers) can be formed by multiplying prime numbers.

@@

\vskip +15 pt
\paragraph*{The number of prime numbers $PI(x)$}
\index{PI(x), $\Pi(x)$}
\mbox{}
\vskip +10 pt
The definition for the number\footnote{%
Often, instead of PI(x) the convention $\Pi(x)$ is used.
} $PI(x)$ is similar: It is the number of all primes
that do not exceed the maximum value $x$:

>>>>

\newpage
% \hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_Primes}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, Jan. 2010, May 2013)

@@

% --------------------------------------------------------------------------
\section{What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $4$ are composite numbers, and can be formed by multiplying prime
numbers.

@@

\vskip +15 pt
\paragraph*{The number of prime numbers $PI(x)$}
\index{PI(x), $\Pi(x)$}
\mbox{}
\vskip +10 pt
The definition for the number\footnote{%
Often, instead of PI(x) the convention $\Pi(x)$ is used.
} $PI(x)$ is similar: It is the number of all primes
that do not exceed the maximum value $x$.
